Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2017-04-01 bn.tex
finetuning + Teemu's comments
Finetuning
small fixes

A Bayesian network $B=(G,\theta)$ defines a probability distribution for
$X$. The component $G$ defines the structure of the model as a
directed acyclic graph that has exactly one node for each component of
$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
the nodes from which there are a directed edges to the variable
$X_{i}$.

@@

It is common to assume the uniform prior for structures, in which case
the objective function for structure learning is reduced to the
marginal likelihood $P(D|G,\alpha)$.  If the model parameters
$\theta_{ij}$ are further assumed to be independently Dirichlet
distributed only depending on $i$ and $G_{i}$ and the data $D$ is
assumed to have no missing values, the marginal likelihood can be
decomposed as
\begin{eqnarray}
\label{eqn:bayesmix}
P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}
P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
\end{eqnarray}
In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
in a closed form~\cite{Bunt91, Heck95}.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates regularity in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML-code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

@@

fNML solves the parameter sensitivity problem and yields predictive
models superior to BDeu.  However, the criterion does not satisfy the
property of giving the same score for models that correspond to the
same dependence statements. Furthermore, the learned structures are
often rather complex (see Figure~\ref{fig:bcnpmean}) which also
hampers their interpretation. The quest for a model selection
criterion that would yield more parsimonious, easier to interpret, but
still predictive Bayesian networks structures is one of the main
motivations for this work.

>>>>

A Bayesian network $B=(G,\theta)$ defines a probability distribution for
$X$. The component $G$ defines the structure of the model as a
directed acyclic graph (DAG) that has exactly one node for each component of
$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
the nodes from which there are a directed edges to the variable
$X_{i}$.

@@

It is common to assume the uniform prior for structures, in which case
the objective function for structure learning is reduced to the
marginal likelihood $P(D|G,\alpha)$.  If the model parameters
$\theta_{ij}$ are further assumed to be independently Dirichlet
distributed only depending on $i$ and $G_{i}$ and the data $D$ is
assumed to have no missing values, the marginal likelihood can be
decomposed as
\begin{eqnarray}
\label{eqn:bayesmix}
\lefteqn{P(D|G,\alpha)}\nonumber\\
&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}P(D[G_i=j,i]|\alpha)\nonumber\\
&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}\int P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.
\end{eqnarray}
In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
in a closed form~\cite{Bunt91, Heck95}.

@@

Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
properties of the BDeu score and showed that in certain settings BDeu
has tendency to add more and more parent variables for a child node
even though the empirical conditional entropy of the child given the
parents has already reached zero. In more detail, assume that in our
data $D$, the values of $X_i$ are completely determined by variables
in set $Z$, so that the empirical entropy $H(X_i | Z) = 0$. Now, if we can further
find one or more variables, denoted by $Y$, whose values are
determined completely by the variables in $Z$, then BDeu will prefer
the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
argues that this kind of behaviour violates regularity in model
selection as the more complex model is preferred over a simpler one
even though it does not fit the data any better. The phenomenon seems
to stem from the way the hyperparameters for the Dirichlet
distribution are chosen in BDeu as using Jeffreys' prior,
$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
from this anomaly. However, using Jeffreys' prior causes marginal
likelihood score not to be score equivalent.

@@

Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the BDeu parameter sensitivity problem by using the
NML-code for the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D';G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values,
\begin{equation}
reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
\end{equation}
While the naive
computation of the regret is still prohibitive, Silander et al.\
approximate it efficiently using a so-called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp1}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}

@@

fNML solves the parameter sensitivity problem and yields predictive
models superior to BDeu.  However, the criterion does not satisfy the
property of giving the same score for models that correspond to the
same dependence statements. The score equivalence is usually viewed desirable when DAGs are considered only as models for conditional independence, without any causal interpretation. Furthermore, the learned structures are
often rather complex (see Figure~\ref{fig:bcnpmean}) which also
hampers their interpretation. The quest for a model selection
criterion that would yield more parsimonious, easier to interpret, but
still predictive Bayesian networks structures is one of the main
motivations for this work.

### 2017-04-01 qnml.tex
small fixes

\begin{theorem}\label{consistency}
Let $r_i$ and $q_i$ denote the number of possible values for variable
$X_i$ and its possible configurations of parents $X_{G_i}$,
respectively. As $N \to \infty$
$$
s^{qNML}_i(D;G) =  \log P(D_i \ | \ \hat{\theta}_{i | G_i} )  - \frac{q_i(r_i - 1)}{2} \log N.
$$
\end{theorem}

>>>>

\begin{theorem}\label{consistency}
Let $r_i$ and $q_i$ denote the number of possible values for variable
$X_i$ and its possible configurations of parents $G_i$,
respectively. As $N \to \infty$
$$
s^{qNML}_i(D;G) =  \log P(D_i \ | \ \hat{\theta}_{i | G_i} )  - \frac{q_i(r_i - 1)}{2} \log N.
$$
\end{theorem}
